### 2019-05-09 papers/haskell19/1-intro.tex
3rd person citation to dragen2
Minor changes based on Nachi's feedback

% --------------------------------------
% Intro
%
Random property-based testing is a powerful technique for finding bugs
\cite{ArtsHNS15, HughesPAN16, HughesNSA16, MidtgaardJKNN17}.
%
In the Haskell community, QuickCheck is the predominant tool for that task
\cite{ClaessenH00}.
%
The developers specify
%
\begin{inparaenum}[(i)]
\item the testing properties their systems must fulfill, and
\item random data generators (or generators for short) for the data types
  involved at their properties.
\end{inparaenum}
%
Then, QuickCheck generates random values using these generators, and uses them
to evaluate the testing properties in search of possible counterexamples, which
often indicate the presence of bugs.

>>>>

% --------------------------------------
% Intro
%
Random property-based testing is a powerful technique for finding bugs
~\cite{ArtsHNS15, HughesPAN16, HughesNSA16, MidtgaardJKNN17}.
%
In Haskell, QuickCheck is the predominant tool for this task \cite{ClaessenH00}.
%
The developers specify
%
\begin{inparaenum}[(i)]
\item the testing properties their systems must fulfill, and
\item random data generators (or generators for short) for the data types
  involved at their properties.
\end{inparaenum}
%
Then, QuickCheck generates random values using these generators, and uses them
to evaluate the testing properties in search of possible counterexamples, which
often indicate the presence of bugs.

### 2019-05-09 papers/haskell19/2-sources.tex
Minor changes based on Nachi's feedback

\begin{code}
simplify :: Html -> Html
simplify (Join (Text t1) (Text t2)) = Text (t1 ++ t2)
simplify (Join (Join (Text t) x) y)
  = simplify (Text t1 iJoin simplify (x iJoin y))
simplify (Join x y) = simplify x iJoin simplify y
simplify (Tag t x) = Tag t (simplify x)
simplify x = x
\end{code}
%
This function simply traverses |Html| values, joining together every contiguous
pair of |Text| constructors.
%
Ideally, we would like to put approximately the same testing effort info each
clause of |simplify|, or perhaps even more to the first two ones, since those
are the ones performing actual simplifications.
%
However, these two clauses are the most difficult ones to test in practice!
%
The probability of generating a random value satisfying nested patterns
decreases multiplicatively with the number of constructors we simultaneously
pattern match against to.

@@

\begin{code}
  gen d = frequency
    [ ...
    , (2, do  t1  <- arbitrary
              t2  <- arbitrary
              return (Text t1 iJoin Text t2))
    , (4, do  t2  <- arbitrary
              x   <- go (d-1)
              y   <- go (d-1)
              return ((Text t1 iJoin x) iJoin y))]
\end{code}%$

>>>>

\begin{code}
simplify :: Html -> Html
simplify (Join (Text t1) (Text t2)) = Text (t1 ++ t2)
simplify (Join (Join (Text t) x) y)
  = simplify (Text t1 iJoin simplify (x iJoin y))
simplify (Join x y) = simplify x iJoin simplify y
simplify (Tag t x) = Tag t (simplify x)
simplify x = x
\end{code}
%
This function simply traverses |Html| values, joining together every contiguous
pair of |Text| constructors.
%
Ideally, we would like to put approximately the same testing effort into each
clause of |simplify|, or perhaps even more to the first two ones, since those
are the ones performing actual simplifications.
%
However, these two clauses are the most difficult ones to test in practice!
%
The probability of generating a random value satisfying nested patterns
decreases multiplicatively with the number of constructors we simultaneously
pattern match against.

@@

\begin{code}
gen d = frequency
  [ ...
  , (2, do  t1  <- arbitrary
            t2  <- arbitrary
            return (Join (Text t1) (Text t2)))
  , (4, do  t2  <- arbitrary
            x   <- go (d-1)
            y   <- go (d-1)
            return (Join (Join (Text t1) x) y))]
\end{code}

### 2019-05-09 papers/haskell19/3-representation.tex
Minor changes based on Nachi's feedback

\begin{code}
class Functor f => Algebra f a | f -> a where
  alg :: f a -> a
\end{code}
%
where |f| is the functor type used to represent a construction of the target
type |a|.
%
Note the functional dependency |f -> a|, helping the type checker to solve
uniquely the free type variable |a| at the right hand side of the |=>|, based on
the type |f| used to represent it.
%
% In practical terms, it ensures that each representation type |f| will be
% translated to a unique target type |a|.
%
Then, we need to instantiate this type class for each data constructor
representation we are considering, providing an appropriate implementation for
the overloaded |alg| function.
%
To do so, we can simply transform each constructor representation by picking its
corresponding data constructor, and piping its fields unchanged:

@@

\begin{code}
eval :: Algebra f a => Fix f -> a
eval = alg . fmap eval . unFix
\end{code}
%
This function exploits the |Functor| structure of our representations,
unwrapping the fixed points and mapping their algebras to the result of
evaluating recursively each recursive sub-term.
%
This recursion schema using functors and F-Algebras to fold data generically has
been well studied and is known a \emph{catamorphism} \cite{Meijer91}.

>>>>

\begin{code}
class Functor f => Algebra f a | f -> a where
  alg :: f a -> a
\end{code}
%
where |f| is the functor type used to represent a construction of the target
type |a|.
%
Note the functional dependency |f -> a|, helping the type checker to solve
uniquely the free type variable |a| on the right hand side of the |=>|, based on
the type |f| used to represent it.
%
% In practical terms, it ensures that each representation type |f| will be
% translated to a unique target type |a|.
%
Then, we need to instantiate this type class for each data constructor
representation we are considering, providing an appropriate implementation for
the overloaded |alg| function.
%
To do so, we can simply transform each constructor representation by picking its
corresponding data constructor, and piping its fields unchanged:

@@

\begin{code}
eval :: Algebra f a => Fix f -> a
eval = alg . fmap eval . unFix
\end{code}
%
This function exploits the |Functor| structure of our representations,
unwrapping the fixed points and mapping their algebras to the result of
evaluating recursively each recursive sub-term.
%
This recursion schema using functors and F-Algebras to fold data generically has
been well studied and is known as a \emph{catamorphism} \cite{Meijer91}.

### 2019-05-09 papers/haskell19/7-related.tex
Minor changes based on Nachi's feedback

From the practical point of view, \citeauthor{KiriyamaOptimizingDTC} propose an
optimization mechanism for Data Types \`a Carte, where a concrete data type has
to be derived for each different composition of constructions defined by the
user \cite{KiriyamaOptimizingDTC}.
%
This solution avoids much of the runtime overhead introduced when internally
pattern matching against sequences of |InL| and |InR| data constructors.
%
However, we see this solution as not entirely compositional, since we still need
to rely on Template Haskell to derive the machinery for each specialized data
type.
%
In our particular setting, we found that our solution has a fairly acceptable
overhead, achieved by automatically balancing our representation types.

>>>>

From the practical point of view, \citeauthor{KiriyamaOptimizingDTC} propose an
optimization mechanism for Data Types \`a la Carte, where a concrete data type
has to be derived for each different composition of constructions defined by the
user \cite{KiriyamaOptimizingDTC}.
%
This solution avoids much of the runtime overhead introduced when internally
pattern matching against sequences of |InL| and |InR| data constructors.
%
However, we see this solution as not entirely compositional, since we still need
to rely on Template Haskell to derive the machinery for each specialized data
type.
%
In our particular setting, we found that our solution has a fairly acceptable
overhead, achieved by automatically balancing our representation types.

### 2019-05-09 papers/haskell19/abstract.tex
Minor changes based on Nachi's feedback

Generating ``good'' random values described by algebraic data types is often
quite intricate.
%
State-of-the-art tools for synthesizing random generators serve the valuable
purpose of helping with this task, while providing different level of invariants
imposed over the generated values.
%
However, none of them are built neither for composability nor extendability, a
useful feature when the shape of our random data makes a difference while
testing different properties or sub-systems.
%
% Instead, users are forced to decide wether to manually write their random
% generators or to use rigid automatically derived ones.

>>>>

Generating ``good'' random values described by algebraic data types is often
quite intricate.
%
State-of-the-art tools for synthesizing random generators serve the valuable
purpose of helping with this task, while providing different level of invariants
imposed over the generated values.
%
However, none of them are built for composability nor extensibility, a useful
feature when the shape of our random data makes a difference while testing
different properties or sub-systems.
%
% Instead, users are forced to decide wether to manually write their random
% generators or to use rigid automatically derived ones.
